---
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    code_folding: hide
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: spacelab
  pdf_document:
      toc: true
  knitrBootstrap::bootstrap_document:
    theme.chooser: TRUE
    highlight.chooser: TRUE
---

```{r setup}
#### Libraries needed
.libPaths(R.home('library'))

suppressPackageStartupMessages(library('ggplot2'))
suppressPackageStartupMessages(library('reshape2'))
suppressPackageStartupMessages(library('rtracklayer'))
suppressPackageStartupMessages(library('genomation'))
suppressPackageStartupMessages(library('scales'))
suppressPackageStartupMessages(library('GenomicRanges'))

## Preset variables

device = "png"
outputIsHTML = TRUE
theme = theme_bw()
digits = 2
tmpdir <- getwd()
theCall = NULL
startTime <- Sys.time()
```

## Calculate observed and reference repeat counts

In this section, observed recovered repeat counts intersecting peaks are summarized.

```{r obs}
rmsk_annotated_peaks<-snakemake@params[["rmsk_annotated_peaks"]]
obslist<-lapply(rmsk_annotated_peaks,import.gff)
names(obslist)<-sub("_rmsk.bed","",basename(rmsk_annotated_peaks))

tablist<-lapply(obslist,function(X){
  a<-strsplit(mcols(X)$gene_id,"/")
  b<-lapply(a,function(Y)Y[[1]])
  table(unlist(b))
})

```

Now we calculate the reference repeat counts from the repeat masker file.

```{r ref}
rmsk_gtf<-snakemake@params[["rmsk_gtf"]]
ref<-import.gff(rmsk_gtf)
r1<-strsplit(mcols(ref)$gene_id,"/")
r2<-lapply(r1,function(Y)Y[[1]])
reftab<-table(unlist(r2))

```

Now, we combine the observe and the reference counts.


```{r comb}
tablist$Genome<-reftab

tablistt<-vector("list",length(tablist))
names(tablistt)<-names(tablist)
  for(i in 1:length(tablist)){X<-as.data.frame(t(t(tablist[[i]])))
          X<-X[,-c(2)]
          colnames(X)<-c("Repeat",names(tablist)[i])
  tablistt[[i]]<-X        
  }
#put ref and obs counts in a table, calculate recovered percentages
tabdat<-Reduce(function(dtf1, dtf2) merge(dtf1, dtf2, by = "Repeat", all = TRUE),tablistt)
tabdatNorm<-apply(tabdat[,!colnames(tabdat) %in% c("Repeat","Genome")],2,function(X)X/tabdat$Genome)
rownames(tabdatNorm)<-tabdat$Repeat
```

Now, we save the calculated table and produce a plot of percentage recovered repeat instances.

```{r melt and plot recovered}
plotdat<-as.data.frame(tabdatNorm)
plotdat$Repeat<-rownames(tabdatNorm)
plotdat2<-melt(plotdat,id.var="Repeat",value.name="FractionRecovered",variable.name="Experiment")
write.table(plotdat2,"FractionRecoveredRepeats.tsv",sep="\t",row.names=FALSE,quote=FALSE)

#select a number of repeat names for plotting
plotdatsub<-plotdat2[plotdat2$Repeat %in% c("Alu","ERV1","ERVK","ERVL","ERVL-MaLR","hAT-Charlie","L1","L2","MIR","TcMar-Tigger"),]
ggplot(data=plotdatsub)+geom_bar(position="stack",aes(x=Experiment,y=FractionRecovered,fill=Repeat),stat="identity")+theme(axis.text.x=element_text(angle=90))
ggsave("FractionRecoveredRepeats_by_Experiment.png")

```

## Calculate repeat counts expected by chance

In this section, we're going to calculate repeat counts expected by chance after peak interval randomization, and compare the observed counts to the expected counts.


```{r exp}
rmsk_annotated_randomized_peaks<-snakemake@params[["rmsk_annotated_randomized_peaks"]]
randlist2<-lapply(rmsk_annotated_randomized_peaks,import.gff)
names(randlist2)<-sub("_rmsk.bed","",basename(rmsk_annotated_randomized_peaks))

randlist3<-lapply(randlist2,function(X){
  a<-strsplit(mcols(X)$gene_id,"/")
  b<-lapply(a,function(Y)Y[[1]])
  table(unlist(b))
})


randlist3$Genome<-reftab

randlist3t<-vector("list",length(randlist3))
names(randlist3t)<-names(randlist3)
  for(i in 1:length(randlist3)){X<-as.data.frame(t(t(randlist3[[i]])))
          X<-X[,-c(2)]
          colnames(X)<-c("Repeat",names(randlist3)[i])
  randlist3t[[i]]<-X        
  }

```


Now, we're going to put refence and expected counts in a table, then calculate percentages of repeat counts recovered by chance.

```{r ref_exp}
randdat<-Reduce(function(dtf1, dtf2) merge(dtf1, dtf2, by = "Repeat", all = TRUE),randlist3t)
randdatNorm<-apply(randdat[,!colnames(randdat) %in% c("Repeat","Genome")],2,function(X)X/randdat$Genome)
rownames(randdatNorm)<-randdat$Repeat

```

Now, we're going to put together observed and expected repeat counts.

```{r obs_exp}
all(as.character(tabdat$Repeat)==as.character(randdat$Repeat)) #FALSE
tabdat$Repeat<-as.character(tabdat$Repeat)
randdat$Repeat<-as.character(randdat$Repeat)
randdat<-randdat[match(tabdat$Repeat,randdat$Repeat),]
randdat_clean<-randdat[!is.na(randdat$Repeat),]
tabdat_clean<-tabdat[!is.na(randdat$Repeat),]
OEdat<-as.matrix(tabdat_clean[,!colnames(tabdat_clean) %in% c("Repeat","Genome")])/as.matrix(randdat_clean[,!colnames(randdat_clean)%in% c("Repeat","Genome")])
rownames(OEdat)<-tabdat_clean$Repeat

l2OEdat<-log2(OEdat)
rownames(l2OEdat)<-rownames(OEdat)

```

Now, we're going to reshape the data, write to a table and plot.

```{r melt and plot obs to exp}
oeplotdat<-OEdat[rownames(OEdat) %in% c("Alu","L1","L2","LTR"),,drop=FALSE]
l2oeplotdat<-l2OEdat[rownames(l2OEdat) %in% c("Alu","L1","L2","LTR"),,drop=FALSE]

l2oeplotdat2<-as.data.frame(l2oeplotdat)
l2oeplotdat2$Repeat<-rownames(l2oeplotdat)
l2oeplotdat3<-melt(l2oeplotdat2,id.var="Repeat",value.name="Log2ObsVsExp",variable.name="Experiment")
write.table(l2oeplotdat3,"Log2Observed_v_Expected.tsv",,sep="\t",row.names=FALSE,quote=FALSE)

ggplot(data=l2oeplotdat3)+geom_dotplot(aes(x=Repeat,y=Experiment,fill=Log2ObsVsExp),position="identity")+scale_fill_gradient2(
  high = muted("red"),
  mid = "white",
  low = muted("blue"),
  midpoint = 0,
  space = "Lab",
  na.value = "grey50",
  transform = "identity",
  guide = "colourbar",
  aesthetics = "fill"
)

ggsave("Log2ObservedVsExpectedRepeats_by_Experiment.png")
```

## Reproducibility

```{r reproducibility1, echo=FALSE}
## Date the report was generated
Sys.time()
```

Wallclock time spent generating the report.

```{r reproducibility2, echo=FALSE}
## Processing time in seconds
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits=3)
```

`R` session information.

```{r reproducibility3, echo=FALSE}
## Session info
options(width = 120)
sessionInfo()
```
