---
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
title: "WGBS DSS QC and Analysis Summarization"
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
.libPaths(R.home("library"))

require("DSS")
require("bsseq")
require("ggplot2")
require("FactoMineR")
require("car")

sampleSheet = snakemake@params[['sampleSheet']]
groups = snakemake@params[['groups']]
##smoothing = snakemake@params[['smoothing']]
smoothing = TRUE
##smoothingSpan = snakemake@params[['smoothingSpan']]
smoothingSpan = 500
maxDist = snakemake@params[['maxDist']]
minCpGs = snakemake@params[['minCpGs']]
minMethDiff = snakemake@params[['minMethDiff']]
minCoverage = snakemake@params[['minCoverage']]
blacklist = snakemake@params[['blacklist']]
FDR = snakemake@params[['FDR']]
BPPARAM = MulticoreParam(workers = snakemake@threads, progressbar=TRUE)

odmls = sprintf("%s/DMLs.txt", snakemake@params[["odir"]])
odmrs = sprintf("%s/DMRs.txt", snakemake@params[["odir"]])
ordata = sprintf("%s/Session.RData", snakemake@params[["odir"]])

ss = read.delim(sampleSheet)
infiles = sprintf("MethylDackel/%s_CpG.bedGraph", ss$name)

g1 = which(ss$condition == groups[1])
g2 = which(ss$condition == groups[2])
```

# Overview

This report summarizes the differentially methylated loci (DML) and regions (DMR) called using DSS of hte samples specified in **`r snakemake@params[["sampleSheet"]]`**.

# DMLs and QC

```{r echo=FALSE, warning=FALSE, message=FALSE}
bs = read.bismark(infiles, colData=ss, BPPARAM=BPPARAM, BACKEND="HDF5Array") # nThread?
sampleNames(bs) = as.character(ss$name)
bptasks(BPPARAM) = 0L  # read.bismark resets this for some reason

if(length(blacklist)) {
    bl = import.bed(blacklist)
    bs = subsetByOverlaps(bs, bl, invert=TRUE)
}

RMV <- which(DelayedMatrixStats::rowSums2(getCoverage(bs, type="Cov") < minCoverage) != 0)
if(length(RMV)) {
    bs = bs[-RMV,]
}

dmlT = DMLtest(bs, group1=g1, group2=g2, smoothing=smoothing, smoothing.span=smoothingSpan, BPPARAM=BPPARAM)
dmls = callDML(dmlT, delta=minMethDiff)
dmls = dmls[which(dmls$fdr < FDR), ]
write.table(dmls, file=odmls, quote=FALSE, sep="\t", row.names=FALSE)
if(nrow(dmlT) > 0) {
    dmrs = callDMR(dmlT, delta=minMethDiff, dis.merge=maxDist, minCG=minCpGs)
} else {
    dmrs = NULL
}
save.image(file=ordata)
write.table(dmrs, file=odmrs, quote=FALSE, sep="\t", row.names=FALSE)

# This ensures that nrow(dmrs) doesn't ever throw an error
dmls = ifelse(is.null(dmls), matrix(NA, nrow=0, ncol=1), dmls)
dmrs = ifelse(is.null(dmrs), matrix(NA, nrow=0, ncol=1), dmrs)
```

Using a minimum methylation difference of `r minMethDiff` and FDR of `r FDR` there are a total of `r nrow(dmls)` DMLs. The DMLs are saved in `r odmls`. The distribution of the methylation changes at these is show below. Note that assymmetric changes in the distribution indicate global shifts in methylation between the groups.

```{r echo=FALSE, warning=FALSE, message=FALSE}
if(nrow(dmls) > 0) {
  g = ggplot(dmlT, aes(x=diff)) + geom_histogram() + labs(x="Methylation Difference (per-CpG)")
  g = g + geom_vline(xintercept=minMethDiff) + geom_vline(xintercept=-1*minMethDiff)
  g
} else {
  message('No DMLs found.')
}
```

Similarly, the test statistic, which is used to compute the p-value, is shown below. Its interpretation is the same as that above.

```{r echo=FALSE, warning=FALSE, message=FALSE}
if(nrow(dmls) > 0) {
  ggplot(dmlT, aes(x=stat)) + geom_histogram() + labs(x="Test statistic (per-CpG)")
} else {
  message('No DMLs found.')
}
```

The p-value distribution for the DMLs should be relatively flat with a spike near 0 and possibly 1. If there's a notable slope or bump elsewhere then there is likely a confounding variable missing from your design (the WGBS pipeline does not currently support complex designs).

```{r echo=FALSE, warning=FALSE, message=FALSE}
if(nrow(dmls) > 0) {
  ggplot(dmlT, aes(x=pval)) + geom_histogram() + labs(x="Unadjusted p-value (per-CpG)")
} else {
  message('No DMLs found.')
}
```

A PCA of the sample is indicated below. It uses the logit-transformed methylation values.

```{r echo=FALSE, warning=FALSE, message=FALSE}
if (nrow(dmls) > 0) {
  x = PCA(logit(getBSseq(bs, type="M") / (1e-3 + getBSseq(bs, type="Cov")), adjust=1e-3), graph=FALSE)
  plot.PCA(x, choix="var")
} else {
  message('No DMLs found.')
}
```

# DMRs and QC

DMR calling functions on the output of DML testing, in essence by looking for neighboring DMLs and merging them together into regions. If we consider a minimum methylation difference of `r minMethDiff`, a maximum distance between CpGs of `r maxDist` and then regions with at least `r minCpGs` CpGs, there are a total of `r nrow(dmrs)` DMRs. These are saved in `r odmrs`. The distribution of DMR lengths is shown below. In general there should be relatively few small (<100 bases) DMRs with a bump in enrichment somewhere between 100 bases and 10kb.

```{r echo=FALSE, warning=FALSE, message=FALSE}
if(nrow(dmrs) > 0) {
  ggplot(dmrs, aes(x=length)) + geom_histogram() + labs(x="DMR length (bases)")
} else {
  message("No DMRs found.")
}
```

There is typically an exponentially decreasing relationship between the number of DMRs found and the number of CpGs in them, which is shown below.

```{r echo=FALSE, warning=FALSE, message=FALSE}
if(nrow(dmrs) > 0) {
  ggplot(dmrs, aes(x=nCG)) + geom_histogram() + labs(x="CpGs per DMR")
} else {
  message("No DMRs found.")
}
```

As with DMLs, there can often be an assymmetry in the distribution of methylation changes in DMRs. Such assymmetries indicate a global shift in methylation changes.

```{r echo=FALSE, warning=FALSE, message=FALSE}
if(nrow(dmrs) > 0) {
  g = ggplot(dmrs, aes(x=diff.Methy)) + geom_histogram() + labs(x="Methylation Difference (per-DMR)")
  g = g + geom_vline(xintercept=minMethDiff) + geom_vline(xintercept=-1*minMethDiff)
  g
} else {
  message("No DMRs found.")
}
```

If there were DMRs, the first 10 (assuming there were that many) ordered by the sum of test statistics.

```{r echo=FALSE, warning=FALSE, message=FALSE}
if(nrow(dmrs) > 0) {
  foo = min(10, nrow(dmrs))
  for(i in 1:foo) {
    print(showOneDMR(dmrs[i,], bs))
  }
} else {
  message("No DMRs found.")
}
```

# Session Information

```{r echo=FALSE, warning=FALSE, message=FALSE}
save.image(file=ordata)
```

This session has been saved to `r ordata` and session information is shown below:

```{r echo=FALSE}
sessionInfo()
```
