---
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
title: "WGBS Metilene QC and Analysis Summarization"
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
.libPaths(R.home("library"))

suppressMessages(require(GenomicRanges))
suppressMessages(require(GenomicFeatures))
suppressMessages(require(dplyr))
suppressMessages(require(tidyr))
suppressMessages(require(ggplot2))

DMRinput = snakemake@input[["DMRs"]]
PCAinput = snakemake@input[["CpGs"]]
gtfFile = snakemake@params[["genes_gtf"]]
FDR = snakemake@params[["FDR"]]
minMethDiff = snakemake@params[["minMethDiff"]]
# Only a single output file is allows for Rmd files
outputTxtFile = sprintf("%s/DMRs.FDR%s.annotated.txt", snakemake@params[["outdir"]], snakemake@params[["FDR"]])
outputPCA = sprintf("%s/PCA.png", snakemake@params[["outdir"]])
outputDiffMethyl = sprintf("%s/MethylationDensity.png", snakemake@params[["outdir"]])
outputQValue = sprintf("%s/QValueDistribution.png", snakemake@params[["outdir"]])
outputVolcano = sprintf("%s/Volcano.png", snakemake@params[["outdir"]])
d = read.delim(DMRinput)
```

# Overview

This report summarizes the differentially methylated region (DMR) calling with metilene of the samples specified in **`r snakemake@params[["sampleSheet"]]`**.

# DMR annotation

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Read in the input data
d = read.delim(DMRinput)
colnames(d)[1:(length(colnames(d))-2)] = c("CHROM", "START", "END", "qvalue", "MeanDiff", "NumCpGs", "pMWU", "p2DKS")
gr = GRanges(seqnames=d$CHROM,
             ranges=IRanges(start=d$START, end=d$END),
             mcols=d[,c(4:10)])
# Fix meta data column names
colnames(elementMetadata(gr)) = gsub("mcols.", "", colnames(elementMetadata(gr)))

# Get the genes
gtf = GenomicFeatures::makeTxDbFromGFF(gtfFile, format="gtf")
g = genes(gtf)

# Subset g to only contain seqlevels in gr
seqlevels(g, pruning.mode="tidy") = intersect(seqlevels(g), seqlevels(gr))

################################
# Annotated DMRs
################################
# Annotate with the nearest gene and its distnace
og = distanceToNearest(gr, g, ignore.strand=FALSE, select="all")
d = t(sapply(split(og, queryHits(og)), function(x) {
      idx = queryHits(x)[1]
      gnames = paste0(g$gene_id[subjectHits(x)], collapse=",")
      distances = paste0(elementMetadata(x)$distance, collapse=",")
      return(as.character(c(idx, gnames, distances)))
}))
d = as.data.frame(d,stringsAsFactors=FALSE)
colnames(d) = c("idx", "NearestGene", "DistanceToNearestGene")
d$idx = as.numeric(d$idx)
gr$NearestGene = as.character("")
gr$DistanceToNearestGene = as.character("")
gr$NearestGene[d$idx] = d$NearestGene
gr$DistanceToNearestGene[d$idx] = d$DistanceToNearestGene

# Write some output
d = as.data.frame(gr)
colnames(d)[1] = c("chrom")

# Filter output by FDR threshold and MeanDiff
IDX = which(d$qvalue < FDR)
write.table(d[IDX,c(1:3, 6:14)], file=outputTxtFile, sep="\t", col.names=T, row.names=F, quote=F)
```

The genome annotation in **`r gtfFile`** was used to annotated DMRs according to their proximity to genes. The results for the **`r length(IDX)`** candidate DMRs passing the FDR and methylation difference threshold are stored in **`r outputTxtFile`**.

# PCA plot

PCA plot calculated on single CpGs used as input to metilene is contained in and can copied directly from this html document. Typically, the first principal component represents the auto-correlation in methylation values for single CpGs. If the biological effect of the experiment on DNA methylation is sufficiently large, the samples should split by biological condition on the second principal component.

```{r echo=FALSE, warning=FALSE, message=FALSE}
d = read.table(PCAinput,header=TRUE,sep="\t",quote="",as.is=FALSE)
d = subset(d,select=!(colnames(d) %in% c("chr","pos")))
x1<-FactoMineR::PCA(d,graph=FALSE)
if(nrow(x1$eig)>=2){
FactoMineR::plot.PCA(x1,choix="var")
}

```

# Distribution of methylation values per group

Averge methylation values for candidate DMRs in each of the groups are plotted below and stored in **`r outputDiffMethyl`**. In general, the starkness of the difference in the distributions should correlate to the number of candiate DMRs.

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Select the mean* columns, strip "mean" and make long
d = as.data.frame(elementMetadata(gr,stringsAsFactor=FALSE)) %>% select(starts_with("mean", ignore.case=FALSE)) %>% 
  rename_all(funs(gsub("^mean_", "", .))) %>% gather("Group", "MeanMethylation")

g = ggplot(d, aes(x=MeanMethylation, fill=Group, color=Group, group=Group)) + geom_density(alpha=0.3)
g = g + theme(text=element_text(size=16),
              axis.text = element_text(size=12),
              axis.title = element_text(size=14))
g = g + xlab("Mean methylation ratio") + xlim(0,100)
g = g + scale_fill_manual(values=c("grey28", "red", "darkblue", "darkgreen"),aesthetics = c("colour", "fill"))
ggsave(outputDiffMethyl, plot=g)
g
```

# Q-value distribution

Q values should ideally show an enrichment around 0 and 1 with an otherwise approximately uniform distribution elsewhere. If you do not see this, then there was likely a batch effect in the data or other QC issues with some of the samples. The vertical line at **`r FDR`** indicates the specified significance threshold. The following plot is saved as **`r outputQValue`**.

```{r echo=FALSE, warnings=FALSE, messages=FALSE}
d = as.data.frame(elementMetadata(gr)) %>% select("qvalue")
g = ggplot(d, aes(x=qvalue)) + geom_histogram(binwidth=0.005)
g = g + theme(text=element_text(size=16),
              axis.text = element_text(size=12),
              axis.title = element_text(size=14))
g = g + xlab("Q value")
g = g + geom_vline(aes(xintercept=as.numeric(FDR)))
ggsave(outputQValue, plot=g)
g
```

# Volcano plot

Volcano plots show the relationship between the effect size (methylation change) and Q value. Typically an increase in the former results in an increase in the significance of the latter. Here, the values are colored according to the FDR (**`r FDR`**) and **`r minMethDiff`**. The following plot is also saved as **`r outputVolcano`**.

```{r echo=FALSE, warning=FALSE, message=FALSE}
d = as.data.frame(elementMetadata(gr))
d$Significant = ifelse(d$qvalue < FDR & abs(d$MeanDiff) > minMethDiff, "Yes", "No")
g = ggplot(d, aes(x=MeanDiff, y=-log10(d$qvalue))) + geom_point(aes(color=Significant))
g = g + theme(text=element_text(size=16),
              axis.text=element_text(size=12),
              axis.title=element_text(size=14))
g = g + labs(x="Mean methylation difference", y="-log10(Q value)")
g = g + scale_color_manual(values=c("grey28", "red", "darkblue", "darkgreen"))
ggsave(outputVolcano, plot=g)
g
```

# Session information

The following packages were used to produce this report:

```{r echo=FALSE}
sessionInfo()
```
