---
title: "WGBS dmrseq QC and Analysis Summarization"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
---

```{r warning=FALSE, message=FALSE, error=FALSE}
.libPaths(R.home("library"))

require("dmrseq")
require("BiocParallel")
require("ggplot2")

sampleSheet = snakemake@params[['sampleSheet']]
groups = snakemake@params[['groups']]

# These aren't actually used yet
maxDist = snakemake@params[['maxDist']]
minCpGs = snakemake@params[['minCpGs']]
minMethDiff = snakemake@params[['minMethDiff']]
minCoverage = snakemake@params[['minCoverage']]
FDR = snakemake@params[['FDR']]
blacklist = snakemake@params[['blacklist']]

BPPARAM = MulticoreParam(workers = snakemake@threads, progressbar=TRUE)

odmrs = sprintf("%s/DMRs.txt", snakemake@params[["odir"]])
ordata = sprintf("%s/Session.RData", snakemake@params[["odir"]])

ss = read.delim(sampleSheet)
row.names(ss) = as.character(ss$name)
infiles = sprintf("MethylDackel/%s_CpG.bedGraph", ss$name)

g1 = which(ss$condition == groups[1])
g2 = which(ss$condition == groups[2])

message(sprintf("Samples belonging to group 1 (%s) are %s",groups[1],ss$name[g1]))
message(sprintf("Samples belonging to group 2 (%s) are %s",groups[2],ss$name[g2]))
```

# Overview

This report summarizes the differentially methylated regions (DMR) called using dmrseq of the samples specified in **`r snakemake@params[["sampleSheet"]]`**.

# DMRs and QC

```{r loadData, warning=FALSE, message=FALSE, error=FALSE, results=FALSE}
bs = read.bismark(infiles, colData=ss, BPPARAM=BPPARAM, BACKEND="HDF5Array", verbose=FALSE)
bptasks(BPPARAM) = 0L  # read.bismark resets this for some reason

# TODO: test this!
if(length(blacklist)) {
    bl = import.bed(blacklist)
    bs = subsetByOverlaps(bs, bl, invert=TRUE)
}

RMV <- which(DelayedMatrixStats::rowSums2(getCoverage(bs, type="Cov") < minCoverage) != 0)
if(length(RMV)) {
    bs = bs[-RMV,]
}

regions = dmrseq(bs=bs, cutoff=0.05, testCovariate = "condition", BPPARAM=BPPARAM, verbose=FALSE)
write.table(regions, file=odmrs, quote=FALSE, sep="\t", row.names=FALSE)
save.image(file=ordata)
```

Using a minimum methylation difference of `r minMethDiff` and FDR of `r FDR` there are a total of `r length(regions)` DMRs. The distribution of the methylation changes at these is shown below. Note that assymmetric changes in the distribution indicate global shifts in methylation between the groups.

```{r warning=FALSE, message=FALSE}
if (length(regions) > 0) {
  g = ggplot(as.data.frame(regions), aes(x=beta)) + geom_histogram() + labs(x="Methylation Difference (per-DMR)")
  g = g + geom_vline(xintercept=minMethDiff) + geom_vline(xintercept=-1*minMethDiff)
  g
} else {
  message('No DMRs found.')
}
```

Similarly, the test statistic, which is used to compute the p-value, is shown below. Its interpretation is the same as that above.

```{r warning=FALSE, message=FALSE}
if (length(regions) > 0) {
  g = ggplot(as.data.frame(regions), aes(x=stat)) + geom_histogram() + labs(x="Methylation Difference Statistic")
  g
} else { 
  message('No DMRs found.')
}
```

It can be useful to plot the unadjusted p-values for the DMRs to help in diagnosing cases where fewer statistically significant DMRs than expected are called. In general, the unadjusted p-value distribution should be fairly uniform, with an increase in counts near 0. Other distributions indicate that there may be confounding effects that should be incorporated in the statistical model.

```{r warning=FALSE, message=FALSE}
if (length(regions) > 0) {
  g = ggplot(as.data.frame(regions), aes(x=pval)) + geom_histogram(binwidth=0.005) + labs(x="Unadjusted p-value")
  print(g)
} else {
  message('No DMRs found.')
}
```


# Session Information

This session has been saved to `r ordata` and session information is shown below:

```{r echo=FALSE}
sessionInfo()
```
