---
title: "WGBS_QC_report"
author: "`r Sys.info()[length(Sys.info())-1]`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
params:
    QCdir: "/data/processing3/sikora/A215.test2/QC_metrics"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Conversion rate

Commercial bisulfite conversion kits claim an efficiency of >98%. In practice, the expected conversion rate can be taken as >95%.
Incomplete bisulfite conversion will result in uninterpretable methylation values.
Bisulfite conversion rate is calculated on non-CpG cytosines in the sequencing reads.

```{r}
knitr::opts_chunk$set(echo = FALSE)
QCdir<-params$QCdir
message(sprintf("Processing quality metrics stored in %s",QCdir))
convRatedir=dir(QCdir,pattern=".conv.rate.txt$",full.names=TRUE)
if(length(convRatedir>0)){
    convRate.short<-gsub(".conv.rate.txt","",basename(convRatedir))

    convRateL<-vector("list",length(convRatedir))
    names(convRateL)<-convRate.short

    for(i in seq_along(convRateL)){
        tabi<-read.table(convRatedir[i],header=FALSE,sep="\t",quote="",as.is=TRUE)
        colnames(tabi)<-c("Read",names(convRateL)[i])
        convRateL[[i]]<-tabi
    }
    convRatedat<-Reduce(function(...) merge(..., all=T,by="Read"), convRateL)
    plottab<-as.data.frame(t(convRatedat),stringsAsFactors=FALSE)[-1,]
    colnames(plottab)<-c("READ1","READ2")
    require(pander)
    pander(plottab,style='grid',caption="Conversion rate in forward and reverse mates.")}
```

## Mapping rate

A good mapping rate of 80-90% indicates successful library construction. Low mapping rates may reflect issues with sample contamination, or damaged starting material.

```{r , echo=FALSE}
QCdir<-params$QCdir
Mapdir=dir(QCdir,pattern="*flagstat$",full.names=TRUE)
Mapdir.short<-gsub(".flagstat","",basename(Mapdir))

bamRepL<-vector("list",length(Mapdir))
names(bamRepL)<-Mapdir.short

for(i in seq_along(bamRepL)){

tot<-system(paste0('grep \"in total\" ',Mapdir[i]),intern=TRUE)
mapped<-system(paste0('grep \"mapped\" ',Mapdir[i]),intern=TRUE)[1]
tabi<-as.data.frame(cbind(as.numeric(unlist(strsplit(tot,split=" "))[1]),as.numeric(unlist(strsplit(mapped,split=" "))[1])))
colnames(tabi)<-c("total","mapped")
tabi$pct<-with(tabi,mapped/total*100)

bamRepL[[i]]<-tabi

}

bamReptab<-as.data.frame(do.call(rbind,bamRepL),stringsAsFactors=FALSE)

require(pander)
pander(bamReptab,style='simple',caption="Mapping rate.")
```

## PCR duplication rate

Observed PCR duplication rate on a PCR-free protocol should be low, not exceeding 10-15%. Higher duplication rates might reflect low-input samples and/or lower library complexity.

```{r , echo=FALSE}
QCdir<-params$QCdir
Mapdir=dir(gsub("QC_metrics","bams",QCdir),pattern="*.rm_dupes.err$",full.names=TRUE,recursive=TRUE)
Mapdir.short<-gsub(".rm_dupes.err","",basename(Mapdir))

bamRepL<-vector("list",length(Mapdir))
names(bamRepL)<-Mapdir.short

for(i in seq_along(bamRepL)){

PE<-system(paste0('grep \"end pairs\" ',Mapdir[i]),intern=TRUE)
SE<-system(paste0('grep \"single ends\" ',Mapdir[i]),intern=TRUE)
dup<-system(paste0('grep \"found\" ',Mapdir[i]),intern=TRUE)
tabi<-as.data.frame(cbind(as.numeric(unlist(strsplit(PE,split=" "))[4]),as.numeric(unlist(strsplit(SE,split=" "))[7]),as.numeric(unlist(strsplit(dup,split=" "))[4])))
colnames(tabi)<-c("PE","SE","Dupes")
tabi$IN_dupes<-with(tabi,2*PE+SE)
tabi$Pct_dupes<-with(tabi,Dupes/IN_dupes*100)

bamRepL[[i]]<-tabi

}

bamReptab<-as.data.frame(do.call(rbind,bamRepL),stringsAsFactors=FALSE)

require(pander)
pander(bamReptab,style='simple',caption="PCR duplication rate.")
```

## Genome-wide coverage

For the purpose of method comparison and publication, genome-wide coverage is calculated with GATK depth of coverage. Reads with MAPQ >= 10 are counted.

```{r,echo=FALSE}
QCdir<-params$QCdir
mq10.intCovdir<-dir(QCdir,pattern="*.mean.genome.doc.sample_summary",full.names=TRUE)
mq10.short<-gsub(".mean.genome.doc.sample_summary","",basename(mq10.intCovdir))

mq10covL<-vector("list",length(mq10.intCovdir))
names(mq10covL)<-mq10.short

for(i in seq_along(mq10covL)){
tabi<-read.table(mq10.intCovdir[i],header=TRUE,sep="\t",quote="",colClasses=c(rep("NULL",2),"numeric",rep("character",3),rep("NULL",4),"numeric",rep("NULL",4)),fill=TRUE)
tabi$SampleID<-names(mq10covL)[i]
mq10covL[[i]]<-tabi[1,,drop=FALSE]
}

mq10covdat<-do.call(rbind,mq10covL)
mq10plottab<-as.data.frame(mq10covdat[,c("mean")],stringsAsFactors=FALSE)
colnames(mq10plottab)<-"MeanCoverage"
rownames(mq10plottab)<-mq10covdat$SampleID
require(pander)
pander(mq10plottab,style='simple',caption="Mean genome-wide coverage.")
```

## CpG dinucleotide coverage

Fraction of CpG dinucleotides covered with at least 10 reads of MAPQ >=10 is calculated on a random sample of 1mln CpGs.

```{r,echo=FALSE}
QCdir<-params$QCdir
mq10.intCovdir<-dir(QCdir,pattern="*.mean.CG.doc.sample_summary",full.names=TRUE)
mq10.short<-gsub("mean.CG.doc.sample_summary","",basename(mq10.intCovdir))

mq10covL<-vector("list",length(mq10.intCovdir))
names(mq10covL)<-mq10.short

for(i in seq_along(mq10covL)){

tabi<-read.table(mq10.intCovdir[i],header=TRUE,sep="\t",quote="",colClasses=c(rep("NULL",2),"numeric",rep("character",3),rep("NULL",4),"numeric",rep("NULL",4)),fill=TRUE)
tabi$SampleID<-names(mq10covL)[i]
mq10covL[[i]]<-tabi[1,,drop=FALSE]

}

mq10covdat<-do.call(rbind,mq10covL)

mq10plottab<-as.data.frame(mq10covdat[,c("X._bases_above_10")],stringsAsFactors=FALSE)
colnames(mq10plottab)<-"FrxnBasesAbove10"
rownames(mq10plottab)<-mq10covdat$SampleID
require(pander)
pander(mq10plottab,style='simple',caption="Fraction of CpG dinucleotides covered at least 10x at MAPQ >=10.")
```

#Methylation bias

Library construction of standard directional BS-Seq samples often consist of several steps including sonication, end-repair, A-tailing and adapter ligation. Since the end-repair step typically uses unmethylated cytosines for the fill-in reaction the filled-in bases will generally appear unmethylated after bisulfite conversion irrespective of their true genomic methylation state. (from sequencing.qcfail.com)
Mbias is obtained using MethylDackel.

```{r , echo=FALSE}
QCdir<-params$QCdir
Mbiasdir<-dir(QCdir,pattern="*.Mbias.txt$",full.names=TRUE)
Mbias.short<-gsub(".Mbias.txt","",basename(Mbiasdir))
message(sprintf("Processing mbias files:  %s",Mbiasdir))

MbiasL<-vector("list",length(Mbiasdir))
names(MbiasL)<-Mbias.short

for(i in seq_along(MbiasL)){

tabi<-read.table(Mbiasdir[i],header=FALSE,sep="-",quote="",colClasses=c("NULL","NULL","character","NULL","character"))
colnames(tabi)<-c("OT","OB")
MbiasL[[i]]<-tabi

}

Mbiasdat<-do.call(rbind,MbiasL)

require(pander)
pander(Mbiasdat,style='simple',caption="Inclusion bounds on top and bottom strand reads.")

```

